\chapter{Number Systems}
\label{chap:ch2}
Numbers are the foundation of mathematics, representing quantities, measurements, and relationships in both abstract and concrete forms. The study of numbers dates back thousands of years, with early civilisations like the Babylonians, Egyptians, and Greeks laying the groundwork for modern number theory. These ancient cultures developed basic arithmetic, including addition, subtraction, multiplication, and division, which are still taught today.

Number theory, often called the "Queen of Mathematics," is a branch of mathematics devoted to the study of integers and the relationships between them. It encompasses a wide range of topics, including prime numbers, divisibility, modular arithmetic, and the properties of number systems. Number theory has a rich history, with significant contributions from mathematicians such as Euclid, who proved the infinitude of prime numbers around 300 BCE, and Pierre de Fermat, known for Fermat's Last Theorem, a problem that puzzled mathematicians for over 350 years until it was finally solved by Andrew Wiles in 1994.

\begin{figure}[htbp]
    \centering
\begin{tikzpicture}[scale=0.7]

    % Complex Numbers (outer square)
    \draw[thick, fill=mseComplex] (-12, -6) rectangle (10, 5);
    \node at (5.5, 4) {\Huge $\mathbb{C}$};
    
    % Real Numbers (largest set)
    \draw[thick, fill=mseReals] (-4,0) ellipse (7.8cm and 4.5cm);
    \node at (-3, 3) {\Large $\mathbb{R}$};
    
    % Rational Numbers
    \draw[thick, fill=mseRationals] (-7,0) ellipse (4.5cm and 3cm);
    \node at (-11, 0) {\Large$\mathbb{Q}$};
    
    % Integers
    \draw[thick, fill=mseIntegers] (-7,0) ellipse (3cm and 2cm);
    \node at (-9.5, 0) {\Large$\mathbb{Z}$};
    
    % Natural Numbers
    \draw[thick, fill=mseNaturals] (-7,0) ellipse (2cm and 1.2cm);
    \node at (-7, 0) {\Large$\mathbb{N}$};
    
    % Irrational Numbers
    \draw[thick, fill=mseIrrationals] (1,0) ellipse (2cm and 1.2cm);
    \node at (1, 0) {\Large$\mathbb{R} \setminus \mathbb{Q} $};
    
    % Imaginary Numbers
    \draw[thick, dashed, fill=mseImaginary] (7,0) ellipse (2cm and 1.2cm);
    \node at (7, 0) {\Large $\mathbb{I}$};

\end{tikzpicture}
    \caption{Venn diagram of numbers}
    \label{fig:venn_num}
\end{figure}

Throughout history, number theory has evolved from a purely theoretical pursuit to a field with practical applications in modern technology. For example, cryptography, the science of securing communication, relies heavily on number theory, particularly the properties of prime numbers and modular arithmetic. The algorithms that protect our online transactions and digital communications are built upon the principles of number theory.

In this chapter, we will explore various aspects of numbers and number theory, starting with the basics of integers, prime numbers, and number systems, and gradually advancing to more complex topics. By understanding the fundamental properties of numbers, we gain insights into the mathematical structures that underpin the digital world and beyond.

\begin{remark}
    \autoref{fig:venn_num} may appear misleading at first glance, as it suggests that there are real numbers which are neither rational nor irrational (depicted by the blue region). However, this is not the case. The set of real numbers is exclusively comprised of rational and irrational numbers. This is precisely why the set of irrational numbers is denoted as \(\mathbb{R} \setminus \mathbb{Q}\), meaning "all real numbers except the rational numbers". The same is true of the complex numbers; the are all composed of the real numbers and the imaginary numbers.
\end{remark}


\begin{table}[ht]
\centering
\caption{Types of Numbers and Their Properties}
\label{tab:number_types}
\renewcommand{\arraystretch}{1.6} % Adjust the row height for the entire table
\begin{tabular}{|m{3.1cm}|>{\centering\arraybackslash}m{1.5cm}|m{5.4cm}|m{5cm}|}
\hline
\raggedright\text{Natural Numbers} & \(\mathbb{N}\) & Numbers used for counting (all positive integers). & \(0, 1, 2, \ldots\) \\ \hline
\raggedright\text{Integers} & \(\mathbb{Z}\) & All positive and negative whole numbers. & \(\{\ldots, -2, -1, 0, 1, 2, \ldots\}\) \\ \hline
\raggedright\text{Rational Numbers} & \(\mathbb{Q}\) & All real numbers which can be expressed as a fraction, \(\frac{p}{q}\), where \(p\) and \(q\) are integers and \(q \neq 0\). All integers are rational numbers as \(1\) is a non-zero integer. & \(\dfrac{1}{5}, \dfrac{5}{1} (=5), \dfrac{2}{3}, \dfrac{3}{2}, \dfrac{0}{3} (=0)\) \\ \hline
\raggedright\text{Irrational Numbers} & \(\mathbb{R}\ \setminus \mathbb{Q}\) & All real numbers which cannot be expressed as a fraction whose numerator and denominator are integers (i.e., all real numbers which aren't rational). & \(\pi, \sqrt{2}, \sqrt{3}\) \\ \hline
\raggedright\text{Real Numbers} & \(\mathbb{R}\) & Includes all numbers on the number line. & \(\dfrac{1}{5}, \sqrt{\dfrac{1}{5}}, 0, -2\) \\ \hline
\raggedright\text{Imaginary Numbers} & \(\mathbb{I}\) & Numbers which are the product of a real number and the imaginary unit \(i\) (where \(i = \sqrt{-1}\)). & \(\begin{array}{l}
\vspace{3pt} 3i = \sqrt{-9},\ -5i = \sqrt{-25}, \\
3\sqrt{2}i = \sqrt{-18} \vspace{3pt}
\end{array}\) \\ \hline
\raggedright\text{Complex Numbers} & \(\mathbb{C}\) & All numbers which can be expressed in the form $a+b i$ where $a$ and $b$ are real numbers and $i=\sqrt{-1}$. Each complex number is a combination of a real number $(a)$ and an imaginary number $(b i)$. & $1+2 i, 1, i,-3 i, 0,-5+i$ \\ \hline
\end{tabular}
\end{table}

\begin{remark}
Many numbers are included in more than one set. \autoref{tab:number_types} offers an overview of the names, properties of and symbols used for the main number types.
\end{remark}

\section{Types of Numbers}
Let's explore the various types of numbers, including natural numbers, whole numbers, integers, rational numbers, irrational numbers, and real numbers - imaginary numbers and complex numbers are left out of this discussion.

\subsection*{Natural Numbers}
We begin with the natural numbers. We distinguish between whole numbers: $0, 1, 2, 3, \ldots$ and counting numbers: $1, 2, 3, \ldots$. These numbers are primarily used for counting. Natural numbers are often regarded as exact values (e.g., there are 4 tires on a car, 8 legs on a spider). However, in some contexts, they may be used as approximations (e.g., there were approximately 1000 people in the crowd).

One of the key properties of natural numbers is that they are \textit{closed} under certain operations, such as addition and multiplication. This means that if you take any two natural numbers and add or multiply them, the result will always be another natural number. For example, if $a$ and $b$ are natural numbers, then both $a + b$ and $a \times b$ are also natural numbers.

There is very little consensus as to whether the symbol $\mathbb{N}$ includes 0. Therefore, the set of whole numbers (that include 0) is often denoted $\mathbb{W}$.

\subsection*{Integers}
Integers are the basic building blocks of number theory, consisting of the set of whole numbers and their negatives. Formally, integers include numbers like $\ldots,-3,-2,-1,0,1,2,3, \ldots$ Basic arithmetic operations, such as addition and multiplication, follow certain properties:

\begin{custombox}{{Commutative, Associative, and Distributive Laws}}


    \begin{itemize}
    \item Commutativity: $a+b=b+a$
    \item Associativity: $(a+b)+c=a+(b+c)$
    \item Distributive property: $a \times(b+c)=a \times b+a \times c$
    \end{itemize}
    
\end{custombox}

Integers are a fundamental set of numbers in mathematics, denoted by the symbol $\mathbb{Z}$. This set includes all the positive and negative whole numbers, as well as zero. Integers extend the natural numbers by incorporating their additive inverses, thereby allowing for the complete operation of subtraction within the set.

\textbf{Additive Identity:} The number \(0\) is known as the additive identity. This is because for any integer \(a\), adding zero does not change the value of \(a\). In mathematical terms, this property is expressed as \(0 + a = a + 0 = a\). This identity is crucial because it ensures that the set of integers remains stable under addition.
    
 \textbf{Multiplicative Identity:} Similarly, the number \(1\) serves as the multiplicative identity. For any integer \(a\), multiplying by one leaves the value of \(a\) unchanged: \(1 \times a = a \times 1 = a\). This property underpins the stability of integers under multiplication, maintaining the integrity of the set.
    
\textbf{Additive Inverse:} For each integer \(a\), there exists a corresponding additive inverse, denoted as \(-a\). The additive inverse is defined such that when \(a\) and \(-a\) are added together, the result is the additive identity, zero: \(a + (-a) = 0\). This property allows for the operation of subtraction within the set of integers, as subtraction can be viewed as the addition of an additive inverse.

\begin{custombox}{Closure Property}
    \textbf{Addition:} The set of integers is closed under the operation of addition. This means that if you take any two integers and add them together, the sum will always be an integer. For example, if \(a\) and \(b\) are integers, then \(a + b\) is also an integer. This closure property ensures that the set of integers is stable and complete under addition, meaning that no matter how many times you add integers together, the result will remain within the set of integers.

\textbf{Multiplication:} The set of integers is also closed under multiplication. If you multiply any two integers, the product will always be an integer. For instance, if \(a\) and \(b\) are integers, then \(a \times b\) is also an integer. This property guarantees that the operation of multiplication, like addition, does not produce results outside the set of integers, thereby preserving the integrity of the set under multiplication.

\end{custombox}

It is universally agreed upon that the definition of an integer is clear and precise. Therefore, when in doubt, it is advisable to refer to numbers within this set as "integers." When you specifically need to refer to only the positive integers, it is both accurate and professional to explicitly state "positive integers." This terminology not only ensures clarity but also reflects a sound understanding of mathematical conventions. 

Remember, zero is neither positive nor negative, so when discussing subsets of integers, careful consideration of this fact is necessary:

\begin{itemize}
    \item \textbf{Integers:} $\mathbb{Z} = \{\ldots,-4,-3,-2,-1,0,1,2,3,4, \ldots\}$
    \item \textbf{Negative Integers:} $\mathbb{Z}^- = \{\ldots,-4,-3,-2,-1\}$
    \item \textbf{Positive Integers:} $\mathbb{Z}^+ = \{1,2,3,4, \ldots\}$
    \item \textbf{Non-Negative Integers:} $\mathbb{Z}_0^+ = \mathbb{Z}_{\geq 0} = \{0,1,2,3,4, \ldots\}$
    \item \textbf{Non-Positive Integers:} $\mathbb{Z}_0^- = \mathbb{Z}_{\leq 0} = \{\ldots,-4,-3,-2,-1, 0\}$
\end{itemize}

This notation should be consistent with standard mathematical conventions.

\subsection*{Rational Numbers}
Rational numbers, denoted by the symbol $\mathbb{Q}$, are numbers that can be expressed as the quotient of two integers, where the numerator is any integer and the denominator is a non-zero integer. Formally, a rational number can be written as $\frac{a}{b}$, where $a \in \mathbb{Z}$ and $b \in \mathbb{Z} \setminus \{0\}$. This set of numbers is fundamental in mathematics as it provides a means to represent fractions and ratios, allowing for a wide range of arithmetic operations, and as such are a generalisation of common fractions which we saw in \autoref{chap:ch1}. They can represent any number that can be written as a finite or repeating decimal.

One of the key properties of rational numbers is that each non-zero rational number has a \textit{multiplicative inverse}. The multiplicative inverse of a rational number $\frac{a}{b}$ is the rational number $\frac{b}{a}$, provided that $a \neq 0$. The importance of the multiplicative inverse lies in the fact that when a number is multiplied by its inverse, the result is the \textit{multiplicative identity}, which is 1. In other words, for any rational number $\frac{a}{b}$, its inverse is denoted by $\left(\frac{a}{b}\right)^{-1} = \frac{b}{a}$, and we have:

\[
\frac{a}{b} \times \frac{b}{a} = \frac{ab}{ba} = 1
\]

This property ensures that rational numbers are closed under multiplication and division (except division by zero), making them a robust and versatile set of numbers for various mathematical operations.

Rational numbers are dense on the number line, meaning that between any two rational numbers, there exists another rational number. This property makes the set of rational numbers particularly important in the study of real numbers, as it allows for the approximation of irrational numbers to any desired degree of accuracy.

\subsection*{Irrational Numbers}
Irrational numbers, denoted by the symbol $\mathbb{R} \setminus \mathbb{Q}$, are real numbers that cannot be expressed as the quotient of two integers. Unlike rational numbers, which have a repeating or terminating decimal expansion, the decimal expansion of an irrational number neither repeats nor terminates. This property makes irrational numbers fundamentally different from rational numbers, as they cannot be precisely represented as fractions.

Some of the most famous examples of irrational numbers include:

\begin{itemize}
    \item \textbf{Pi} ($\pi$): Perhaps the most well-known irrational number, $\pi$ is the ratio of the circumference of a circle to its diameter. Its value is approximately $3.14159\ldots$, but its decimal expansion goes on infinitely without repeating. $\pi$ plays a crucial role in geometry, trigonometry, and calculus.

    \item \textbf{The Golden Ratio} ($\phi$): The golden ratio, $\phi \approx 1.61803\ldots$, is an irrational number that appears frequently in nature, art, and architecture. It is defined as the positive solution to the equation $x^2 - x - 1 = 0$, and it is the limit of the ratio of successive Fibonacci numbers.

    \item \textbf{The Square Root of 2} ($\sqrt{2}$): The square root of 2, approximately $1.41421\ldots$, is the length of the diagonal of a square with side length 1. This number is historically significant because its discovery by the ancient Greeks, particularly the Pythagoreans, revealed the existence of numbers that could not be expressed as the ratio of two integers. The Pythagorean theorem states that in a right triangle, the square of the hypotenuse is equal to the sum of the squares of the other two sides. For a right triangle with both legs of length 1, the hypotenuse is $\sqrt{2}$, demonstrating that $\sqrt{2}$ cannot be a rational number.

    \item \textbf{Euler's Number} ($e$): The number $e \approx 2.71828\ldots$ is the base of the natural logarithm and is a fundamental constant in mathematics, particularly in calculus and complex analysis. The number $e$ arises naturally in various growth processes, such as compound interest and population growth.
\end{itemize}

Irrational numbers fill the gaps between rational numbers on the number line, making the real numbers a continuous set. However, like rational numbers, they too have important properties. For instance, while irrational numbers do not have a simple fractional representation, they are nonetheless essential in representing the lengths, areas, and volumes that cannot be captured by rational numbers alone.

\subsection*{Real Numbers}
Real numbers, denoted by the symbol $\mathbb{R}$, form the foundation of most mathematical analysis and are essential in describing continuous quantities. The set of real numbers is composed of both rational numbers ($\mathbb{Q}$) and irrational numbers ($\mathbb{R} \setminus \mathbb{Q}$), thus encompassing all numbers that can be placed on the number line. While both rational and irrational numbers are part of the real number system, they differ significantly in their properties:

\begin{custombox}{Similarities and Differences of Rational and Irrational Numbers}
    \begin{itemize}
    \item \textbf{Representation:} Rational numbers can be represented as fractions, while irrational numbers cannot. This distinction makes irrational numbers more complex to handle, especially in arithmetic operations.

    \item \textbf{Decimal Expansion:} The decimal expansion of rational numbers is either finite or periodic, while that of irrational numbers is infinite and non-repeating.

    \item \textbf{Closure Properties:} Rational numbers are closed under addition, subtraction, multiplication, and division (except by zero). Irrational numbers are not closed under these operations; for example, the sum or product of two irrational numbers can sometimes be rational.

    \item \textbf{Density:} Rational numbers are dense on the real number line, meaning that between any two rational numbers, there exists another rational number. Irrational numbers are also densely distributed but in a complementary manner, filling in the "gaps" left by the rationals, ensuring that the real number line is continuous without any breaks.
\end{itemize}
\end{custombox}

The real number line is a continuous, unbroken line that extends infinitely in both directions. Every point on this line corresponds to a unique real number, whether rational or irrational. This continuity is what allows the real numbers to model continuous phenomena in nature, such as time, distance, and temperature.

\section{Numeral Systems}

We are so accustomed to working within the decimal system that we often forget it is a relatively recent invention and was once considered revolutionary. It is time to carefully examine how we represent numbers. Typically, we use the decimal system, where a number like 3459 is shorthand for \(3 \times 1000 + 4 \times 100 + 5 \times 10 + 9\). The position of each digit is crucial, as it allows us to distinguish between values like 30 and 3. The decimal system is a\textbf{ positional numeral system}, meaning it has designated positions for units, tens, hundreds, and so forth. Each digit’s position implies the multiplier (a power of ten) that should be used with that digit, and each position has a value ten times that of the position to its right.

Notice that we can save space by writing 1000 as \(10^3\), where the exponent 3 indicates the number of zeros. Thus, \(100000 = 10^5\). If the exponent is negative, it represents a fraction, e.g., \(10^{-3} = \frac{1}{1000}\). Perhaps the most ingenious aspect of the positional system was the addition of the decimal point, which allows us to include decimal fractions. For example, the number 123.456 is equivalent to:
\[
1 \times 100 + 2 \times 10 + 3 \times 1 + 4 \times \frac{1}{10} + 5 \times \frac{1}{100} + 6 \times \frac{1}{1000}.
\]

This can be visualised as:
\[
\begin{array}{rccccccccc}
\text{Multiplier:} & \ldots & 10^2 & 10^1 & 10^0 & . & 10^{-1} & 10^{-2} & 10^{-3} & \ldots \\
\text{Digits:} & \ldots & 1 & 2 & 3 & . & 4 & 5 & 6 &\ldots \\
& & & & & \uparrow & & & & \\
& & & & & \text{Decimal Point:} & & & &
\end{array}
\]
However, there is no inherent reason why we must use powers of 10, or base 10. The Babylonians, for instance, used base 60, and base 12 was very common in medieval Europe. Today, the most widely used numeral systems are summarised in \autoref{tab:number_systems}

\begin{table}[ht]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Numeral system} & \textbf{Symbols} & \textbf{Base} & \textbf{Additional information} \\ \hline
\textbf{Decimal} & 0-9 & 10 & - \\ \hline
\textbf{Binary} & 0, 1 & 2 & - \\ \hline
\textbf{Hexadecimal} & 0-9, A-F & 16 & $\mathrm{A} \equiv 10, \mathrm{B} \equiv 11, \mathrm{C} \equiv 12,$ $\mathrm{D} \equiv 13, \mathrm{E} \equiv 14, \mathrm{F} \equiv 15$ \\ \hline
\textbf{Octal} & 0-7 & 8 & - \\ \hline
\end{tabular}
\caption{Summary of Common Numeral Systems}
\label{tab:number_systems}
\end{table}

We begin by focusing on binary which will also receive the most detailed attention in this chapter.

\section{Binary Numbers}
In the binary scale, we express numbers in powers of 2 rather than the 10s of the decimal scale. For some numbers, this is easy. Recall $2^0=1$,

\begin{table}[!ht]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|c c c c c c c c|}
\hline 
\multirow{2}{*}{Decimal number} & & \multirow{2}{*}{In powers of 2} & \multicolumn{4}{c}{Power of 2} & \multirow{2}{*}{Binary number} \\
& & & 3 & 2 & 1 & 0 & \\ \hline
8 & $=$ & $2^3$ & 1 & 0 & 0 & 0 & 1000 \\
7 & $=$ & $2^2 + 2^1 + 2^0$ & 0 & 1 & 1 & 1 & 111 \\
6 & $=$ & $2^2 + 2^1$ & 0 & 1 & 1 & 0 & 110 \\
5 & $=$ & $2^2 + 2^0$ & 0 & 1 & 0 & 1 & 101 \\
4 & $=$ & $2^2$ & 0 & 1 & 0 & 0 & 100 \\
3 & $=$ & $2^1 + 2^0$ & 0 & 0 & 1 & 1 & 11 \\
2 & $=$ & $2^1$ & 0 & 0 & 1 & 0 & 10 \\
1 & $=$ & $2^0$ & 0 & 0 & 0 & 1 & 1 \\ \hline
\end{tabular}
\caption{Decimal Numbers in Binary Representation}
\end{table}

As in decimal, we write this with the position of the digit representing the power, the first place after the decimal being the $2^0$ position, the next the $2^1$, and so on. To convert a decimal number to binary, we can use the \texttt{mod} operator.

As an example, consider 88 in decimal or $88_{10}$. We would like to write it as a binary number. We take the number and successively divide \texttt{mod} 2. See below:

\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|c|c|c|c|c|}
\hline 
Step Number $n$ & $x_n$ & $x_n / 2$ & $x_n \bmod 2$ \\ \hline
0 & 88 & 44 & 0 \\ 
1 & 44 & 22 & 0 \\
2 & 22 & 11 & 0 \\
3 & 11 & 5 & 1 \\
4 & 5 & 2 & 1 \\
5 & 2 & 1 & 0 \\
6 & 1 & 0 & 1 \\ \hline
\end{tabular}
\caption{Conversion of Decimal 88 to Binary}
\end{table}

Writing the last column in reverse, that is from the bottom up, we have 1011000, which is the binary form of 88, i.e., $88_{10} = 1011000_2$.

Binary decimals are less common but quite possible. Thus, 101.1011 is just $2^2 + 2^0 + 2^{-1} + 2^{-3} + 2^{-4}$, which is, after some calculation, 5.6875. We have seen how to turn the integer part of a decimal number into a binary number, and we can do the same with a decimal fraction. Consider 0.6875. As before, we draw up a table:

\begin{table}[h!]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{|c|c|c|c|}
\hline 
Step Number $n$ & $x_n$ & $x_n \times 2$ & $\left\lfloor x_n \times 2 \right\rfloor$ \\ \hline
0 & 0.6875 & 1.375 & 1 \\ \hline
1 & 0.375 & 0.75 & 0 \\ \hline
2 & 0.75 & 1.5 & 1 \\ \hline
3 & 0.5 & 1 & 1 \\ \hline
\end{tabular}
\caption{Conversion of Decimal Fraction 0.6875 to Binary}
\end{table}

Giving, reading down, $0.6875_{10} = 1011_2$.

\subsection*{Binary Expansion}

The process outlined in the previous section is called \textbf{binary expansion} and refers to the representation of a number in the binary (base-2) numeral system. Every decimal number can be expressed as a sum of powers of 2, where each power corresponds to a binary digit (bit) in the number's binary form.

Let's reconsider the decimal number 88. To find its binary expansion, we identify the largest power of 2 less than or equal to 88 and continue subtracting powers of 2 until we reach 0.

First, we note that $2^6 = 64$ is the largest power of 2 less than 88:
\[
88 = 64 + 24
\]

Next, we find that $2^4 = 16$ is the largest power of 2 less than 24:
\[
24 = 16 + 8
\]

Finally, $2^3 = 8$ exactly matches the remainder:
\[
8 = 8 + 0
\]

Thus, we have:
\[
88 = 2^6 + 2^4 + 2^3
\]

In binary, each of these powers of 2 is represented by a '1' in the corresponding place value, with '0' in place values where no power of 2 contributes:
\[
88_{10} = 1011000_2
\]

To summarise:
\begin{itemize}
    \item $2^6 = 64$ corresponds to the leftmost '1' in the binary expansion.
    \item $2^4 = 16$ corresponds to the next '1'.
    \item $2^3 = 8$ corresponds to the next '1'.
    \item The remaining digits are '0' because $2^5$, $2^2$, $2^1$, and $2^0$ do not contribute to the value 88.
\end{itemize}

Thus, the binary expansion of 88 is $1011000_2$. This method of representing numbers is fundamental in computer science and digital electronics, where binary representation is the standard for data storage and processing.

\subsection*{Binary Operations}
Binary operations are basic arithmetic operations performed on binary numbers. These operations are essential in computing and digital systems, as they form the foundation for how computers process and manipulate data.

Binary addition, subtraction, and multiplication are similar to their decimal counterparts but follow simpler rules due to the binary system's limited digits. For example, binary addition follows these rules:

\[
0 + 0 = 0, \quad 0 + 1 = 1, \quad 1 + 0 = 1, \quad 1 + 1 = 10
\]

In this case, \(1 + 1\) results in \(10_2\), which means 0 with a carry of 1 to the next higher bit. Binary subtraction and multiplication follow similar straightforward rules that are easy to implement in digital systems.

The XOR (exclusive OR) operation is another important binary operation. XOR produces a 1 if the two bits being compared are different and a 0 if they are the same:

\[
0 \oplus 0 = 0, \quad 0 \oplus 1 = 1, \quad 1 \oplus 0 = 1, \quad 1 \oplus 1 = 0
\]

In binary addition, the XOR operation is used to add two bits without considering any carry from a previous bit. This is because XOR effectively performs addition modulo 2, which aligns perfectly with how binary addition works. For example:

\[
\begin{array}{c|c|c|c}
\text{Bit 1} & \text{Bit 2} & \text{XOR (Sum)} & \text{AND (Carry)} \\
\hline
0 & 0 & 0 & 0 \\
0 & 1 & 1 & 0 \\
1 & 0 & 1 & 0 \\
1 & 1 & 0 & 1 \\
\end{array}
\]

In the case of \(1 + 1\), XOR gives a sum of 0 and an AND operation (which detects the carry) gives a carry of 1, resulting in the binary number 10.

\begin{align*}
0 + 0 &= 0 \\
0 + 1 &= 1 \\
1 + 1 &= 10 \quad \text{so we carry 1 and leave a zero} \\
1 + 1 + 1 &= 1 + (1 + 1) = 1 + 10 = 11.
\end{align*}

We can write this in very much the same way as for a decimal addition:

\[
\begin{array}{r|r|r|r|r|r|r|r}
& 1 & 1 & 0 & 1 & 0 & 1 & \\
+ & 1 & 0 & 1 & 1 & 1 & 0 & \\
\hline
1 & 1 & 0 & 0 & 0 & 1 & 1 & \quad \text{Sum} \\
\uparrow & & & & \uparrow & & & \\
& & & & & & &
\end{array}
\]

The right-hand arrow shows where we carry a 1. The left-hand arrow shows where we have $1 + 1 + 1$ so we carry a 1 and have a 1 left over.

As we will see below, we will often need to handle multiple carries. There are two ways to handle this which resemble the methods we know from the decimal system. We will explain using an example.

\subsubsection*{Method 1: Column-wise Binary Addition with Multiple Carries}
Consider

\begin{equation*}
\begin{tikzpicture}[
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  &  &  &  &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-4-2.south west) -- (m-4-13.south east);

\end{tikzpicture}
\end{equation*}

\textbf{Step 1: Add the Rightmost Column }\newline
Start by adding the rightmost bits:

$$
1+1+1+1=100_2 \quad(\text { which is binary for } 4)
$$
Reading the result from right to left (i.e. from \textit{least significant bit} (LSB) to the \textit{most significant bit}(MSB))
\begin{itemize}
    \item write down the 0
    \item carry the 0 to the next column
    \item carry the 1 to the third column
\end{itemize}
You end up with

\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  &  & 1 & 0 &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  &  &  & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-5-2.south west) -- (m-5-13.south east);

\end{tikzpicture}
\end{equation*}

\textbf{Step 2: Add the Second Column from the Right}\newline
Next, add the second column:



$$
0+1+0+0+1=10_2 \quad(\text { which is binary for } 2)
$$
Reading the result from LSB to MSB:
\begin{itemize}
    \item write down the 0
    \item carry the 1 to the next column
\end{itemize}
You end up with

\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  &  & 1 &  &   &            \\
        &   &   &   &   &   &  &  &  &  & 1 &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  &  & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-6-2.south west) -- (m-6-13.south east);

\end{tikzpicture}
\end{equation*}


\textbf{Step 3: Add the Third Column from the Right}\newline
Now, add the third column:

$$
1+1+1+1+1+1=110_2 \quad(\text { which is binary for } 6)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 0
    \item carry the 1 to the fourth column
    \item carry the 1 to the fifth column
\end{itemize}
You end up with

Our sum so far:

\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  & 1 & 1 &  &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  & 0 & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-5-2.south west) -- (m-5-13.south east);

\end{tikzpicture}
\end{equation*}


\textbf{Step 4: Add the Fourth Column from the Right}\newline
Move to the fourth column:

$$
1+1+1+1+1=101_2 \quad(\text { which is binary for } 5)
$$

Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 1
    \item carry the 0 to the fifth column
    \item carry the 1 to the sixth column
\end{itemize}
You end up with

\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  & 0 &  &  &  &   &            \\
        &   &   &   &   &   &  & 1 & 1 &  &  &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  & 1 & 0 & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-6-2.south west) -- (m-6-13.south east);

\end{tikzpicture}
\end{equation*}

\textbf{Step 5: Add the Leftmost Column}\newline
Add the leftmost column:

$$
1+1+1+1+1=101_2 \quad(\text { which is binary for } 5)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 1
    \item carry the 0 to the sixth column
    \item carry the 1 to the seventh column
\end{itemize}
This results in

\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  & 0 &  &  &  &  &   &            \\
        &   &   &   &   &   & 1 & 1 &  &  &  &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 0 & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-6-2.south west) -- (m-6-13.south east);

\end{tikzpicture}
\end{equation*}


\textbf{Step 6: Add the Remaining Carries}\newline
Finally, add the remaining carries:

\begin{equation*}
\begin{tikzpicture}[
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  & 1 & 1 & 1 & 1 & 0 & 0 &  0&            \\                                                  
    };

    \draw[-,color=black,semithick] (m-4-2.south west) -- (m-4-13.south east);

\end{tikzpicture}
\end{equation*}

The following example demonstrates the entire process by using different colors to distinguish each column and the corresponding carries they produce. Note that the last two digits in the sum are colored black, as they do not result from any specific column but are instead generated solely from the carries.

\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 3/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  \textcolor{YellowOrange}{1} &  \textcolor{YellowOrange}{0} &  &  &  &  &   &            \\
        &   &   &   &   &   &  & \textcolor{Purple}{1} & \textcolor{Purple}{0} &  & \textcolor{red}{1}  &  &   &            \\
        &   &   &   &   &   &  &  & \textcolor{Green}{1}  & \textcolor{Green}{1}  & \textcolor{blue}{1}  & \textcolor{blue}{0}  &   &            \\
        &  &  &  &  &  &  &  & \textcolor{YellowOrange}{1} & \textcolor{Purple}{1} & \textcolor{Green}{1}  & \textcolor{red}{1}  & \textcolor{blue}{1} &     \\
    +   &  &  &  &  &  &  &  &  \textcolor{YellowOrange}{1} & \textcolor{Purple}{1} & \textcolor{Green}{1}  & \textcolor{red}{0} & \textcolor{blue}{1}  &            \\
    +   &  &  &  &  &  &  &  &  \textcolor{YellowOrange}{1} & \textcolor{Purple}{1} & \textcolor{Green}{1}  & \textcolor{red}{0}  & \textcolor{blue}{1}  &            \\
    +   &  &  &  &  &  &  &  &  \textcolor{YellowOrange}{1} & \textcolor{Purple}{1} & \textcolor{Green}{1}  & \textcolor{red}{1}  & \textcolor{blue}{1}  &            \\
        &  &  &  &  &  & 1 & 1 &  \textcolor{YellowOrange}{1} & \textcolor{Purple}{1} & \textcolor{Green}{0}  & \textcolor{red}{0}  & \textcolor{blue}{1}  &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-7-2.south west) -- (m-7-13.south east);

\end{tikzpicture}
\end{equation*}

\subsubsection*{Method 2: Direct Summation and Simplification}
We will illustrate the second method using the same example. In the previous case, we carried the actual binary number to the next columns. In this method, we write down 0 if the sum is even and 1 if the sum is odd. Every time a sum a multiple of 2, we carry a 1 to the next columns, and then continue this process for each column, including the carries in the calculation of that column.

\textbf{Step 1: Add the Rightmost Column}\newline
Add bits in column 1 (from counting from MSB):

$$
1+1+1+1=100_2 \quad(\text { which is binary for } 4)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 0
    \item carry a 1 for the first multiple of 2
    \item carry a 1 for the second multiple of 2
\end{itemize}
This results in


\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 3/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  &  &  &  &   &            \\
        &   &   &   &   &   &  &  &  &  &  & 1 &   &            \\
        &   &   &   &   &   &  &  & &  &  & 1 &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  &  &  & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-7-2.south west) -- (m-7-13.south east);

\end{tikzpicture}
\end{equation*}

\textbf{Step 2: Add the Second Column from the Right}\newline
Add bits in column 2 (from counting from MSB):

$$
1+1+1+1=100_2 \quad(\text { which is binary for } 4)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 0
    \item carry a 1 for the first multiple of 2
    \item carry a 1 for the second multiple of 2
\end{itemize}
This results in


\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 3/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  &  &  &  &   &            \\
        &   &   &   &   &   &  &  &  &  & 1  &  &   &            \\
        &   &   &   &   &   &  &  & &  & 1 &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  &  & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-7-2.south west) -- (m-7-13.south east);

\end{tikzpicture}
\end{equation*}

\textbf{Step 3: Add the Third Column from the Right}
Add bits in column 3 (from counting from MSB):

$$
1+1+1+1+1+1=110_2 \quad(\text { which is binary for } 6)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 0
    \item carry a 1 for the first multiple of 2
    \item carry a 1 for the second multiple of 2
    \item carry a 1 for the third multiple of 2
\end{itemize}

This results in


\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 3/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  & 1 &  &  &   &            \\
        &   &   &   &   &   &  &  &  & 1 &   &  &   &            \\
        &   &   &   &   &   &  &  & & 1 &  &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  &  & 0 & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-7-2.south west) -- (m-7-13.south east);

\end{tikzpicture}
\end{equation*}

\textbf{Step 4: Add the Fourth Column from the Right}\newline
Add bits in column 4 (from counting from MSB):

$$
1+1+1+1+1+1+1=111_2 \quad(\text { which is binary for } 7)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the 1
    \item carry a 1 for the first multiple of 2
    \item carry a 1 for the second multiple of 2
    \item carry a 1 for the third multiple of 2
\end{itemize}
This results in


\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 3/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  & 1 &  &  &  &   &            \\
        &   &   &   &   &   &  &  & 1 &  &   &  &   &            \\
        &   &   &   &   &   &  &  & 1 &  &  &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  &  &  &  & 1 & 0 & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-7-2.south west) -- (m-7-13.south east);

\end{tikzpicture}
\end{equation*}


\textbf{Step 5: Add the Leftmost Column}\newline
Add bits in leftmost column):

$$
1+1+1+1+1+1+1=111_2 \quad(\text { which is binary for } 7)
$$
Reading the result from LSB to MSB
\begin{itemize}
    \item write down the result in binary, i.e. 1 1 1
\end{itemize}
This results in


\begin{equation*}
\begin{tikzpicture}[
    row 1/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 2/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    row 3/.style={font=\textsl,font=\tiny, anchor=mid,
        inner sep=1.5pt},
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  &  &  &  &   &            \\
        &   &   &   &   &   &  &  &  &  &   &  &   &            \\
        &   &   &   &   &   &  &  & &  &  &  &   &            \\
        &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &     \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 0 & 1 &            \\
    +   &  &  &  &  &  &  &  & 1 & 1 & 1 & 1 & 1 &            \\
        &  &  &  &  &  & 1 & 1 & 1 & 1 & 0 & 0 & 0 &            \\                                                  
    };

    \draw[-,color=black,semithick] (m-7-2.south west) -- (m-7-13.south east);
\end{tikzpicture}
\end{equation*}

which corresponds to the result we obtained above. While not demonstrated explicitly here, subtraction works in a similar fashion.

By using one of these methods for handling multiple carries, allow us to also multiply two binary numbers.



\subsection*{Multiplication in Binary}
Multiplication in binary is technically easier than multiplication in decimal. In binary operations, we work exclusively with two digits: 0 and 1. This means that both the the multiplier\footnote{The "multiplicand" is the number that has to be multiplied, and the "multiplier" is the number by which it is multiplied.} and multiplicand consist of 0's and 1' (and so does the multiplicand). The process of finding the binary product is analogous to traditional multiplication in the decimal system. The four five steps involved in multiplying binary digits are:

\[
\begin{aligned}
& 0 \times 0=0 \\
& 0 \times 1=0 \\
& 1 \times 0=0 \\
& 1 \times 1=1 \\
& 1 \times 10_2 = 10_2 \quad \left( \text{multiplying by base } 10_2 \text{ adds a 0 to the end} \right)
\end{aligned}
\]

The last step means that $101_2 \times 10_2 = 1010_2$ which is analogous to the decimal case: $143_{10} \times 10_{10} = 1430_{10}$.

We will illustrate the process by supplying a couple of examples.

\begin{example}


\begin{equation*}
\begin{tikzpicture}[
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &  (1) &   &   &   &  &  &  &  & 1 & 0 & 0  &            \\
     \times   &   & (2)  &   &   &   &  &  &  &  &   & 1 &  1 &            \\
        &   & (3)  &   &   &   &  &  & &  & 1 & 0 & 0  &            \\
       + &  & (4) &  &  &  &  &  &  & 1 & 0 & 0 & 0 &     \\
        &  & (5) &  &  &  &  &  &  & 1 & 1 & 0 & 0 &            \\                                                  
    };

    \draw[dashed, color=black, semithick] (m-2-2.south west) -- (m-2-13.south east);
    \draw[-,color=black,semithick] (m-4-2.south west) -- (m-4-13.south east);
    \end{tikzpicture}
\end{equation*}

Here are the steps:

\begin{itemize}
    \item Multiply the multiplicand (line 1) by the LSB of the multiplier (line 2), which in this case is 1.
    \item Record this result in line 3.
    \item Append a 0 to line 4 to account for the shift to the next power of 2 in the multiplier.
    \item Multiply the multiplicand (line 1) by the next bit of the multiplier (line 2), which is also 1 in this case.
    \item Add this result to line 4, after the 0 you appended earlier.
    \item Finally, sum the values in lines 3 and 4, as outlined in the previous section, to obtain the final result in line 5.
\end{itemize}
\end{example}

We offer two additional examples:

\begin{example}

\begin{equation*}
\begin{tikzpicture}[
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   &  &  &  &  & 1 & 0 & 1  &            \\
     \times   &   &   &   &   &   &  &  &  & 1 &  0 & 1 &  1 &            \\
        &   &   &   &   &   &  &  & &  & 1 & 0 & 1  &            \\
       + &  &  &  &  &  &  &  &  & 1 & 0 & 1 & 0 &     \\
       + &  &  &  &  &  &  &  & 0 & 0 & 0 & 0 & 0 &            \\               
       + &  &  &  &  &  &  & 1 & 0 & 1 & 0 & 0 & 0 &     \\
        &  &  &  &  &  &  & 1 & 1 & 0 & 1 & 1 & 1 &            \\                                        
    };

    \draw[dashed, color=black, semithick] (m-2-2.south west) -- (m-2-13.south east);
    \draw[-,color=black,semithick] (m-6-2.south west) -- (m-6-13.south east);
    \end{tikzpicture}
\end{equation*}
    
\end{example}

\begin{example}
\label{ex1.3}
\begin{equation*}
\begin{tikzpicture}[
    every node/.style={column sep=.5mm,row sep=1mm}]
    \matrix (m) [matrix of math nodes,
        nodes in empty cells,
        %nodes=draw
    ] 
    {
        &   &   &   &   &   & \textcolor{Green}{1} & \textcolor{Green}{0} & \textcolor{Green}{0} & \textcolor{Green}{1} & \textcolor{Green}{1} & \textcolor{Green}{1} & \textcolor{Green}{0}  &            \\
     \times   &   &   &   &   &   &  &  &  &  &  1 & 0 &  1 &            \\
        &   &   &   &   &   & \textcolor{Green}{1} & \textcolor{Green}{0} & \textcolor{Green}{0} & \textcolor{Green}{1} & \textcolor{Green}{1} & \textcolor{Green}{1} & \textcolor{Green}{0}   &            \\
       + &  &  &  & \textcolor{Green}{1} & \textcolor{Green}{0} & \textcolor{Green}{0} & \textcolor{Green}{1} & \textcolor{Green}{1} & \textcolor{Green}{1} & \textcolor{Green}{0} & \textcolor{Red}{0} & \textcolor{Red}{0} &            \\             
        &  &  &  & 1 & 1 & 0 & 0 & 0 & 0 & 1 & 1 & 0 &     \\                                     
    };

    \draw[dashed, color=black, semithick] (m-2-2.south west) -- (m-2-13.south east);
    \draw[-,color=black,semithick] (m-4-2.south west) -- (m-4-13.south east);
    \end{tikzpicture}
\end{equation*}
    
\end{example}

In \autoref{ex1.3} notice that we omitted the row of zeroes that the second value of the multiplier would have produced, and notice even further that we added two 0's before restating the multiplicand in the sum.

Binary multiplication, like binary addition, is a core operation in computer arithmetic. By breaking the process down into manageable steps—multiplying individual bits and then summing the results—it becomes clear how similar it is to the multiplication methods we use in the decimal system. The main difference is the simplicity and efficiency of working within the binary system, where only the digits 0 and 1 are involved.

We notice how binary multiplication builds on binary addition. Each step, involving shifts and sums, essentially consists of repeated additions adjusted by powers of two. A strong grasp of binary addition naturally leads to a better understanding of binary multiplication and its applications.

\section{Octal and Hexadecimal}
Octal is a base-8 numbering system that uses the digits 0 through 7. It is closely related to binary, which is a base-2 system. The connection between the two lies in how easily binary numbers can be converted to octal and vice versa. Each octal digit corresponds to exactly three binary digits (bits), making conversions straightforward. For example, the binary number `110` converts directly to the octal digit `6`. Because of this close relationship, octal is often used as a shorthand for binary in computing, particularly in contexts where grouping binary digits in sets of three simplifies reading and interpreting binary data.

\[
\begin{aligned}
12_8 &= 1 \cdot 8^1 + 2 \cdot 8^0  = 10_{10} \\
3021_8 & = 3 \cdot 8^3 + 0 \cdot 8^2 + 2 \cdot 8^1 + 1 \cdot 8^0 = 1553_{10}
\end{aligned}
\]

Since \(8\) is \(2^3\), we can express it in binary:

\[
\begin{aligned}
3 & \rightarrow \textcolor{Green}{011} \\
0 & \rightarrow \textcolor{YellowOrange}{000} \\
2 & \rightarrow \textcolor{blue}{010} \\
1 & \rightarrow \textcolor{red}{001} \\
\end{aligned}
\]

\[
\text{Thus, } 3021_8 = \textcolor{Green}{011}\textcolor{YellowOrange}{000} \textcolor{blue}{010}\textcolor{red}{001}_2 = \textcolor{Green}{11}\textcolor{YellowOrange}{000} \textcolor{blue}{010}\textcolor{red}{001}_2
\]

We obtain the final result by removing leading zeros.

Hexadecimal is a base-16 numbering system that uses sixteen distinct symbols: the digits 0-9 and the letters A-F, where A represents 10, B represents 11, and so on up to F, which represents 15. Hexadecimal is closely related to binary because each hexadecimal digit corresponds exactly to four binary digits (bits). This direct relationship makes it easy to convert between the two systems. For example, the hexadecimal digit `A` translates to the binary sequence `1010`. Due to this efficiency in grouping, hexadecimal is often used in computing as a more compact and readable way to represent binary data, particularly in areas like memory addresses and colour codes in web design.

\[
\begin{aligned}
123_{16} &= 1 \cdot 16^2 + 2 \cdot 16^1 + 3 \cdot 16^0  = 256 + 32 + 3 = 291_{10} \\
A2E_{16} & = 10 \cdot 16^2 + 2 \cdot 16^1 + 14 \cdot 16^0 = 2560 + 32 + 14 = 2606_{10}
\end{aligned}
\]

Since \(16\) is \(2^4\), we can, for instance, express $A2E_{16}$ in binary:

\[
\begin{aligned}
A & \rightarrow 1010 \\
2 & \rightarrow 0010 \\
E & \rightarrow 1110 \\
\end{aligned}
\]

\[
\text{Thus, } A2E_{16} = 101000101110_2
\]

Similarly we get $5EB52_{16}$ as

\[
\begin{aligned}
5 & \rightarrow 0101 \\
E & \rightarrow 1110 \\
B & \rightarrow 1011 \\
5 & \rightarrow 0101 \\
2 & \rightarrow 0010 \\
\end{aligned}
\]

\[
\text{Thus, } 5EB52_{16} = 01011110101101010010_2
\]

Again, notice that we removed the leading 0's from $5_{16}$ when writing the result.

\section{Converting Between Systems}

Understanding how to convert numbers between binary, decimal, octal, and hexadecimal systems is essential in computer science and digital electronics. Each system is a different base, and each has its own applications. Here's a step-by-step guide to help you convert numbers from one system to another.

\subsection*{Decimal to Binary Conversion}
To convert a decimal number to binary:
\begin{enumerate}
    \item Divide the decimal number by 2.
    \item Record the remainder (it will be 0 or 1).
    \item Divide the quotient by 2 and record the remainder.
    \item Repeat until the quotient is 0.
    \item The binary number is the sequence of remainders read from bottom to top.
\end{enumerate}

\begin{example} Convert \(23_{10}\) to binary.

\begin{solution}

\[
\begin{aligned}
23 \div 2 & = 11 \quad \text{remainder } 1 \\
11 \div 2 & = 5 \quad \text{remainder } 1 \\
5 \div 2 & = 2 \quad \text{remainder } 1 \\
2 \div 2 & = 1 \quad \text{remainder } 0 \\
1 \div 2 & = 0 \quad \text{remainder } 1 \\
\end{aligned}
\]
Thus, \(23_{10} = 10111_2\). \end{solution}
\end{example}

\subsection*{Decimal to Octal Conversion}
To convert a decimal number to octal:
\begin{enumerate}
    \item Divide the decimal number by 8.
    \item Record the remainder.
    \item Divide the quotient by 8 and record the remainder.
    \item Repeat until the quotient is 0.
    \item The octal number is the sequence of remainders read from bottom to top.
\end{enumerate}

\begin{example}Convert \(78_{10}\) to octal.

\begin{solution}
    


\[
\begin{aligned}
78 \div 8 & = 9 \quad \text{remainder } 6 \\
9 \div 8 & = 1 \quad \text{remainder } 1 \\
1 \div 8 & = 0 \quad \text{remainder } 1 \\
\end{aligned}
\]
Thus, \(78_{10} = 116_8\). \end{solution}
    
\end{example}

\subsection*{Decimal to Hexadecimal Conversion}
To convert a decimal number to hexadecimal:
\begin{enumerate}
    \item Divide the decimal number by 16.
    \item Record the remainder (use A, B, C, D, E, F for remainders 10, 11, 12, 13, 14, 15 respectively).
    \item Divide the quotient by 16 and record the remainder.
    \item Repeat until the quotient is 0.
    \item The hexadecimal number is the sequence of remainders read from bottom to top.
\end{enumerate}

\begin{example}Convert \(255_{10}\) to hexadecimal.

\begin{solution}
    
\[
\begin{aligned}
255 \div 16 & = 15 \quad \text{remainder } 15 \quad (\text{F}) \\
15 \div 16 & = 0 \quad \text{remainder } 15 \quad (\text{F}) \\
\end{aligned}
\]
Thus, \(255_{10} = FF_{16}\).\end{solution}

\end{example}


\subsection*{Binary to Decimal Conversion}
To convert a binary number to decimal:
\begin{enumerate}
    \item Multiply each bit by 2 raised to the power of its position, starting from 0 on the right.
    \item Sum all the products.
\end{enumerate}

Notice that this amounts to the method outlined above about binary expansion.

\begin{example} Convert \(1101_2\) to decimal.

\begin{solution}
    

\[
\begin{aligned}
1 \cdot 2^3 + 1 \cdot 2^2 + 0 \cdot 2^1 + 1 \cdot 2^0 & = 8 + 4 + 0 + 1 = 13_{10}
\end{aligned}
\] \end{solution}

\end{example}

\subsection*{Binary to Octal Conversion}
To convert a binary number to octal:
\begin{enumerate}
    \item Group the binary digits into sets of three, starting from the right. Add leading zeros if necessary.
    \item Convert each group of three binary digits to its octal equivalent.
\end{enumerate}

\begin{example}Convert \(110110_2\) to octal.

\begin{solution}
\[
\begin{aligned}
110 & \rightarrow 6 \\
110 & \rightarrow 6 \\
\end{aligned}
\]
Thus, \(110110_2 = 66_8\). \end{solution}

\end{example}

\subsection*{Binary to Hexadecimal Conversion}
To convert a binary number to hexadecimal:
\begin{enumerate}
    \item Group the binary digits into sets of four, starting from the right. Add leading zeros if necessary.
    \item Convert each group of four binary digits to its hexadecimal equivalent.
\end{enumerate}

\begin{example}Convert \(10110101_2\) to hexadecimal.

\begin{solution}
\[
\begin{aligned}
1011 & \rightarrow B \\
0101 & \rightarrow 5 \\
\end{aligned}
\]
Thus, \(10110101_2 = B5_{16}\). \end{solution}

\end{example}

\subsection*{Octal to Binary Conversion}
To convert an octal number to binary: Convert each octal digit to its 3-bit binary equivalent!

\begin{example}Convert \(57_8\) to binary.

\begin{solution}
\[
\begin{aligned}
5 & \rightarrow 101 \\
7 & \rightarrow 111 \\
\end{aligned}
\]
Thus, \(57_8 = 101111_2\). \end{solution}

\end{example}

\subsection*{Octal to Decimal Conversion}
To convert an octal number to decimal:
\begin{enumerate}
    \item Multiply each digit by 8 raised to the power of its position, starting from 0 on the right.
    \item Sum all the products.
\end{enumerate}

\begin{example}Convert \(157_8\) to decimal.

\begin{solution}
    
\[
\begin{aligned}
1 \cdot 8^2 + 5 \cdot 8^1 + 7 \cdot 8^0 & = 64 + 40 + 7 = 111_{10}
\end{aligned}
\] \end{solution}

\end{example}

\subsection*{Octal to Hexadecimal Conversion}
To convert an octal number to hexadecimal:
\begin{enumerate}
    \item First, convert the octal number to binary.
    \item Then, convert the binary number to hexadecimal by grouping the binary digits in sets of four.
\end{enumerate}

\begin{example}Convert \(157_8\) to hexadecimal.

\begin{solution}

\[
\begin{aligned}
1 & \rightarrow 001 \\
5 & \rightarrow 101 \\
7 & \rightarrow 111 \\
\end{aligned}
\]
Thus, \(157_8 = 001101111_2 = 6F_{16}\). \end{solution}
\end{example}

\subsection*{Hexadecimal to Binary Conversion}
To convert a hexadecimal number to binary: Convert each hexadecimal digit to its 4-bit binary equivalent.

\begin{example}Convert \(2B_{16}\) to binary.

\begin{solution}
    


\[
\begin{aligned}
2 & \rightarrow 0010 \\
B & \rightarrow 1011 \\
\end{aligned}
\]
Thus, \(2B_{16} = 00101011_2\).\end{solution}

\end{example}

\subsection*{Hexadecimal to Decimal Conversion}
To convert a hexadecimal number to decimal:
\begin{enumerate}
    \item Multiply each digit by 16 raised to the power of its position, starting from 0 on the right.
    \item Sum all the products.
\end{enumerate}

\begin{example}Convert \(2B_{16}\) to decimal.

\begin{solution}
\[
\begin{aligned}
2 \cdot 16^1 + 11 \cdot 16^0 & = 32 + 11 = 43_{10}
\end{aligned}
\] \end{solution}
\end{example}

\subsection*{Hexadecimal to Octal Conversion}
To convert a hexadecimal number to octal:
\begin{enumerate}
    \item First, convert the hexadecimal number to binary.
    \item Then, convert the binary number to octal by grouping the binary digits in sets of three.
\end{enumerate}

\begin{example}Convert \(2B_16\) to octal.

\begin{solution}

\[
\begin{aligned}
2 & \rightarrow 0010 \\
B & \rightarrow 1011 \\
\end{aligned}
\]
Thus, \(2B_{16} = 00101011_2 = 53_8\). \end{solution}

\end{example}

\subsection*{Final Thoughts on Conversion}
The concept of expansion plays a central role in these conversions. Whether you are expanding a decimal number into its binary, octal, or hexadecimal form, or converting a binary number into its octal or hexadecimal equivalent, you are more or less expressing the number in terms of powers of the base. The expansion method is essentially the same for each system as it boils down to dividing by the highest power of the base recursively:

$$
\begin{aligned}
7562_{10} & =1 \cdot 16^3+3466=1 \cdot 16^3+13 \cdot 16^2+138 \\
& =\textcolor{red}{1} \cdot 16^3+ \textcolor{blue}{13} \cdot 16^2+\textcolor{YellowOrange}{8} \cdot 16^1+\textcolor{Green}{10} \cdot 16^0=\textcolor{red}{1} \, \textcolor{blue}{D} \, \textcolor{YellowOrange}{8} \, \textcolor{Green}{A}
\end{aligned}
$$

By understanding these expansions and the relationships between these number systems, you can efficiently switch between them, allowing you to represent and manipulate data in the most suitable format for any given situation.
